---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- badges: start -->
[![R-CMD-check](https://github.com/jhuwit/actiwalkability/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jhuwit/actiwalkability/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/jhuwit/actiwalkability/branch/main/graph/badge.svg)](https://codecov.io/gh/jhuwit/actiwalkability?branch=main)
<!-- badges: end -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  cache = TRUE
)
```

# actiwalkability

`actiwalkability` provides helpers for querying the EPA Walkability Index and
working with Census GEOID/FIPS identifiers.

Core entry points:

- `ww_epa_walkability()` for querying the EPA Walkability Index ArcGIS layer
- `ww_fips15()` and `ww_fips12()` for composing Census FIPS/GEOID strings

## Installation

You can install `actiwalkability` from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("remotes")
remotes::install_github("jhuwit/actiwalkability")
```

## Quick start

The EPA index is reported for Census block groups.  Start with an address
lookup that contains 12-character, Census 2010 block-group GEOIDs; keep these
identifiers as character values so leading zeroes are retained.

```{r addresses}
library(actiwalkability)

addresses <- dplyr::tibble(
  address_id = c("address_1", "address_2", "address_3"),
  geoid = c("240054519002", "240054026041", "245102303002")
)

knitr::kable(addresses)
```

If state, county, tract, and block identifiers are available instead, create a
block-group GEOID with `acti_fips12()`:

```{r make-geoid}
acti_fips12(state = 24, county = 510, tract = 60400, block = 2002)
```

Query the EPA layer with the unique GEOIDs.  This is a live service request;
the chunk is cached so subsequent README renders reuse the saved result.

```{r epa-walkability, cache = TRUE}
walkability <- acti_epa_walkability(
  unique(addresses$geoid),
  geometry = FALSE,
  fields = c("GEOID10", "NatWalkInd")
)
```

```{r show-walkability}
knitr::kable(dplyr::select(
  as.data.frame(walkability),
  GEOID10, NatWalkInd, cat_walk_index
))
```

Join the returned index values back to the address lookup:

```{r join-walkability}
address_walkability <- dplyr::left_join(
  addresses,
  walkability,
  by = c("geoid" = "GEOID10")
)

knitr::kable(dplyr::select(
  address_walkability,
  address_id, geoid, NatWalkInd, cat_walk_index
))
```

### Map the example block groups

The EPA values are joined to Census TIGERweb 2010 block-group boundaries, the
boundary vintage that matches `GEOID10`.  The hard-coded plot extent provides
the wider Baltimore, Maryland context, while keeping the README dependencies
small.  This query is cached in the README.

```{r walkability-map, cache = TRUE, fig.width = 7, fig.height = 5}
census_block_groups <- arcgislayers::arc_open(
  "https://tigerweb.geo.census.gov/arcgis/rest/services/Census2020/Tracts_Blocks/MapServer/5"
)
census_where <- paste0(
  "GEOID IN (",
  paste0("'", unique(addresses$geoid), "'", collapse = ", "),
  ")"
)
block_groups_2010 <- arcgislayers::arc_select(
  census_block_groups,
  where = census_where,
  geometry = TRUE,
  fields = c("GEOID", "INTPTLAT", "INTPTLON")
) |>
  dplyr::rename(GEOID10 = GEOID)

walkability_map <- dplyr::left_join(
  block_groups_2010,
  dplyr::select(as.data.frame(walkability), GEOID10, NatWalkInd),
  by = "GEOID10"
)

walkability_labels <- dplyr::mutate(
  as.data.frame(walkability_map),
  longitude = as.numeric(INTPTLON),
  latitude = as.numeric(INTPTLAT)
)

ggplot2::ggplot() +
  ggplot2::geom_sf(
    data = walkability_map,
    ggplot2::aes(fill = NatWalkInd),
    colour = "white"
  ) +
  ggrepel::geom_text_repel(
    data = walkability_labels,
    ggplot2::aes(longitude, latitude, label = GEOID10),
    seed = 2026,
    min.segment.length = 0,
    size = 3
  ) +
  ggplot2::scale_fill_viridis_c(name = "Walkability\nindex") +
  ggplot2::coord_sf(
    crs = 4326,
    default_crs = 4326,
    xlim = c(-77, -76.1),
    ylim = c(39, 39.7),
    expand = FALSE
  ) +
  ggplot2::labs(
    title = "Example Census block groups in Baltimore, Maryland",
    subtitle = "EPA National Walkability Index"
  ) +
  ggplot2::theme_minimal()
```
